Summer 2025 ToDo

### 2025-10-24 Relations/Paper/Summer_2025_TODO.tex

\section*{October 15th TODO}
\begin{todolist}     
  \item It's possible to have a term rewrite system with good properties like FB and the relation being decidable. But not satisfying accCor. That is something that can be added in a final touch of the paper. Either a remark or example, where a term rewrite system does not validate accCor. (In the lambda calculus, there are computable perpetual strategies. We think they're concerned with weak normalization. Not sure if it can be done with strong normalization.)
  \item Do a final read through of the information below the figure. Have we captured all the most important information / Are we repeating ourselves?
  \item Do we want to keep anything in the old/delete section of WF?
  \item Do we want to add anything about counter examples or positive examples? 
  \item We may want to use words rather than macros in certain locations
  \item Check through the comments in general in the section. 
  \item 
\end{todolist}

>>>>

\section*{October 15th TODO / Oct 22nd}
\begin{todolist}     
  \item It's possible to have a term rewrite system with good properties like FB and the relation being decidable. But not satisfying accCor. That is something that can be added in a final touch of the paper. Either a remark or example, where a term rewrite system does not validate accCor. (In the lambda calculus, there are computable perpetual strategies. We think they're concerned with weak normalization. Not sure if it can be done with strong normalization.)
  \item Do a final read through of the information below the figure. Have we captured all the most important information / Are we repeating ourselves?
  \item Do we want to keep anything in the old/delete section of WF?
  \item Do we want to add anything about counter examples or positive examples? 
  \item We may want to use words rather than macros in certain locations
  \item Check through the comments in general in the section. 
  \item (Long term) Investigate: what we've done for natural numbers could be done for an arbitrary ARS (in the counter examples files). The sentence: ''The restriction to $\lnot\lnot$-closed subsets implies (and is equivalent to) 
every subset $P$ being merely \emph{weakly} decidable (satisfying $\lnot \lnot P(x) \lor \lnot P(X)$ 
for all $x$)'' needs further exploration. We could not say it/say it differently. 
\item CHange WFseq not not to WFseq- in the macro. We are sticking with this for now. If we have time we can investigate the relation between the two.
\item Make the diagram horizontal rather than vertical. Spoof it up and we can compare next week.
\item Use of macros ni WF section. Double check the right one is being used. Don't use the "is'' macro at all in this section. Be consistent on whether we are using macros. 
\item 
\end{todolist}

\section*{Plan for PLDI submission}
\begin{todolist}
  \item Aim for first draft finished by the end of the 29th.
  \item Make a new branch at this point to conform to this conference's style?
  \item Submission deadline is Thursday 13th 
  \item Once we've got a current draft finished. Look in Teresa for first order term rewrite systems and see how many we can represent using our code.
\end{todolist}

### 2025-10-24 Relations/Paper/WF_potential_update.tex

\begin{tikzpicture}[
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

% Row 1: WFacc
\node[box] (WFacc) {WFacc};
% \node[box, right=of WFacc] (NNWFacc) {$\lnot\lnot$WFacc};
\node[box, right=2.9cm of WFacc] (WFaccM) {WFacc-};

@@

\draw[arrow] (WFminDNE) to node[left, text = red] {accDNE} (WFacc);
\draw[arrow, bend right=45] (WFminDNE) to node[pos=0.5, left, text=red] {MP$\equiv$} (WFseq);
\draw[arrow, bend right] (WFminDNEM) to node[right, text=red] {FB} (WFaccM);
\draw[arrow, bend right] (WFseqM) to (WFminCor);
\draw[arrow, bend right] (WFminCor) to node [right, text=red] {MP$\equiv$} (WFseqM);

\end{tikzpicture}

>>>>

\begin{figure}[ht]
\centering  
\begin{tikzpicture}[
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

% Row 1: WFacc
\node[box] (WFacc) {$WFacc$};
% \node[box, right=of WFacc] (NNWFacc) {$\lnot\lnot$WFacc};
\node[box, right=2.9cm of WFacc] (WFaccM) {$WFacc\lnot\lnot$};

@@

\draw[arrow] (WFminDNE) to node[left, text = red] {accDNE} (WFacc);
\draw[arrow, bend right=45] (WFminDNE) to node[pos=0.5, left, text=red] {MP$\equiv$} (WFseq);
\draw[arrow, bend right] (WFminDNEM) to node[right, text=red] {FB} (WFaccM);
\draw[arrow, bend right] (WFseqM) to (WFminCor);
\draw[arrow, bend right] (WFminCor) to node [right, text=red] {MP$\equiv$} (WFseqM);
\draw[arrow] (WFCor) -- (WFminCor);
\draw[arrow, bend left = 110] (WFCor) to node[left, text=red] {AccCor} (WFacc);
\draw[arrow, bend left = 80, color=blue] (WFmin) to node[right, text=blue, pos=0.75] {corDNE} (WFCor);

\end{tikzpicture}
\caption{The logical relationships between notions of well-foundedness}
 \label{fig:WF} 
\end{figure}
\clearpage

### 2025-10-24 Relations/Paper/Well-foundedness.tex

% Arrows within rows
\draw[arrow] (WFmin) -- (WFminDNE);
\draw[arrow, bend left=45] (WFmin) to (WFseq);

\draw[arrow] (WFminM) -- (WFminDNEM);
\draw[arrow, bend right=45] (WFminM) to (WFaccM);
\draw[arrow] (WFaccM) -- (WFseqM);
\draw[arrow, bend right=45] (WFaccM) to (WFminDNEM);
\draw[arrow, bend right=45] (WFminDNEM) to (WFminM);
\draw[arrow, bend right=45] (WFminDNEM) to (WFseqM);
\draw[arrow] (WFseqM) -- (WFminCor);

@@

% Additional labeled arrows
\draw[arrow] (WFminDNEM) to node[above, text=red] {$\FB$} (WFaccM);
\draw[arrow, bend right=45] (WFminCor) to node[above, text=red] {$\mpe$} (WFseqM);
\draw[arrow, bend right=45] (WFCor) to node[above, text=red] {$\accCor$} (WFacc);
\draw[arrow] (WFminDNE) to node[below, text=red]{$\accDNE$} (WFacc);
\draw[arrow, bend left = 45] (WFminDNE) to node[below, text=red]{$\mpe$} (WFseq);
\draw[arrow, bend right = 45] (WFminCor) to node[right, text=red]{$\corDNE$} (WFCor);
\draw[arrow, bend right = 45] (WFaccM) to node[right, text=red]{$\accDNE$} (WFacc);
\draw[arrow] (WFacc) to node[below, text=red]{$\rdec$} (WFseq);
\draw[arrow, bend left=45] (WFminCor) to node[below, text=red]{$\accCor$} (WFaccM);

@@

Other observations that can be gleaned from the figure include the following.
\begin{enumerate}
  \item The implication from $\WFacc$ to $\WFseq$ requires the underlying relation to be decidable.
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item $\WFmin$ is a very strong assumption and implies $\WFseq$ unconditionally.
  The more reasonable $\WFminDNE$ requires the image of every sequence to be $\lnot\lnot$-closed
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (if one assumes that equality on $A$ is decidable).
  \item On the $\lnot\lnot$-side, $\WFminm$ and $\WFminDNEm$ are equivalent. Both are directly implied by $\WFaccm$,
    and all three directly imply $\WFseqm$.
  \item All definitions imply $\WFcorm$.  This weakest definition implies $\WFaccm$ assuming the
    ``constructive contrapositive of accessibility'', $\accCor$.
    %While this condition is implied by $\isWFseqm$, it does not appear to imply anything on its own.
  \item Assuming $\accCor$, $\WFacc$ can also be derived from $\WFcor$. % , the stronger version of $\WFcorm$.

@@

  \item At the same time, if the relation is finitely branching ($\FB$), then $\WFminDNEm$
    implies $WFaccm$ directly.

@@

We also did not include the inductive definition, Definition \ref{def:WFproperties}.\ref{def:WFind}.
  Logically, it is trivially equivalent to the accessible one, but its Agda formulation has a defect:
  the inductive well-foundedness property lands in the larger universe $\mathtt{Set}_1$, due to
  quantification over all predicates on $A$.  (This defect is also present in $\WFmin$, $\WFminDNE$, and \texttt{WFcor}.)

@@

\subsection{Old. Delete?}

Translating this statement directly into type theory yields the formula

\[\tag{$\isWFseqm$} \lnot \big(\exists s : \nat \to A \forall k : \nat. \  R \,(s\,(k+1)) \, (s\, k) \big) \]

(The label $\isWFseqm$ refers to the name of the corresponding term in \texttt{Wellfoundedness.agda}.)
Constructively, this is not a useful hypothesis, as it does not provide any positive computational content and requires an explicit sequence to be defined. \sacomment{This sentence added into where we discuss wfseq- as being the textbook definition. Can now be deleted here.}

A stronger formulation asserts that, for any sequence, there exists an element
in the sequence that does not reduce to the next one.  This yields the statement
\[\tag{$\isWFseq$} \forall s : \nat \to A \exists k : \nat.\, \lnot \, R \,(s\,k)\,(s\,(k+1)) \]

These two conditions are classically equivalent, but not constructively so.
Indeed, $\isWFseq$ certainly implies $\isWFseqm$, but the converse only holds classically. \sacomment{We say this through the figure and so can delete it?}

Another classical formulation of well-foundedness
asserts that every non-empty subset
contains a minimal element.  That is,
\[ \tag{$\isWFmin$} \forall P \subseteq A. P \neq \emptyset \to \exists x \in P. (\forall y. Ryx \to y\notin P) \]

While classically equivalent to the previous one, see \cite[Exercise A.1.8]{Terese},
constructively we only have the implication $\isWFmin \then \isWFseq$. \sacomment{Again, this is covered by our new definitions and figure and can be deleted.}

The canonical constructive definition of well-foundedness is that
every inductive predicate is universally true.
This is constructively equivalent to the statement that every element
is accessible.  In Agda, these concepts are encoded as follows.
{\small
\begin{verbatim}
  _-inductive_ : 𝓟 A → Set
  _-inductive_ φ = ∀ x → (∀ y → R y x → φ y) → φ x

  _isWFind : Set₁
  _isWFind = ∀ (φ : 𝓟 A) → _-inductive_ φ → ∀ x → φ x

  data _-accessible : 𝓟 A where
    acc : ∀ {x : A} → (∀ y → R y x → _-accessible y) → _-accessible x

  _isWFacc : Set
  _isWFacc = ∀ (x : A) → x ∈ _-accessible
\end{verbatim}
}
Although the two notions are equivalent, the latter definition allows us to do induction on the proof of accessibility of an individual element $x \in A$.  This makes such proofs easier to write in Agda, thus we take
\verb|_isWFacc| as \emph{the} standard notion of well-foundedness for relations.  \sacomment{We don't talk about which proof is easier to write in Agda, but I don't think we need to include this discussion. I suggest we delete.}

\subsection{Relations between the main definitions}

We have now seen five notions of a well-founded relation, only two of which
are seen to be logically equivalent.

This raises several questions.
\begin{itemize}
\item Are all these notions truly distinct?  If one cannot furnish a proof of implication,
can we provide a counterexample?
\item Under what hypotheses do the implications become provable?
\item How ``admissible'' are such hypotheses, and how often can we hope for them to be valid?
\end{itemize}

The situation of separating these notions is complicated by the fact that, for all
\emph{definable} predicates, the classical and constructive notions do coincide.
(We have not proved this, but see \cite{Berardi} for a related result.)
So there is no hope to discriminate these notions via explicit counterexamples.
Rather, we could attempt to show that one does not imply the other by proving that
such an implication would yield a ``constructive taboo'': a proposition known to be
unprovable without additional classical assumptions.

For example, the archetype of a well-founded relation is the strict order \verb|<|
of natural numbers.  The usual induction readily validates $\WFacc$ for \verb|<|.
  At the same time, the file \texttt{WFCounters.agda} shows that
asserting $\WFmin$ for \verb|<| implies that every subset of the natural numbers is decidable.  Since this is clearly not provable in Agda without classical axioms,
this argument shows that the implication from $\WFacc$ to $\WFmin$ is not provable.

Such distinctions have consequences for rewriting theory.
For example, $x\in \SN_R$ iff $x$ is $\bar{R}$-accessible (where $\bar{R}$ denotes
the converse of $R$).  Thus, if $\bar{R}$ is well-founded, $R \models \gSN$.
Yet actually computing the normal form of a given $x$ seems to require a
strong form of decidability of being $R$-minimal, namely:
\[
\tag{$\isMinDec$} \forall x. \left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)
\]
(Note that asking whether $x$ is a normal form constitutes the weaker disjunction
$\lnot\lnot\left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)$.)

@@

% \greyout{From the perspective of rewriting theory, it would seem wild indeed if
% $\SN$ did not imply $\WN$.
%
These questions motivated us to take a closer look at the issue of well-foundedness.
Our analysis revealed a number of additional related hypotheses, having a complex
network of relations.  %These are displayed in Figure \ref{f:wf}.
These are recorded in the file \verb|Wellfounded.agda| /sacomment{Which file do we now want to refer to here. Or just refer to the entire subfolder?}

>>>>

% Arrows within rows
\draw[arrow] (WFmin) -- (WFminDNE);
\draw[arrow, bend left=25] (WFmin) to (WFseq);

\draw[arrow] (WFminM) -- (WFminDNEM);
\draw[arrow] (WFaccM) -- (WFseqM);
\draw[arrow, bend right=35] (WFaccM) to (WFminDNEM);
\draw[arrow, bend right=35] (WFminDNEM) to (WFminM);
\draw[arrow, bend right=25] (WFminDNEM) to (WFseqM);
\draw[arrow] (WFseqM) -- (WFminCor);

@@

% Additional labeled arrows
\draw[arrow] (WFminDNEM) to node[above, text=red] {$\FB$} (WFaccM);
\draw[arrow, bend right=35] (WFminCor) to node[above, text=red] {$\mpe$} (WFseqM);
\draw[arrow, bend right=25] (WFCor) to node[above, text=red] {$\accCor$} (WFacc);
\draw[arrow] (WFminDNE) to node[below, text=red]{$\accDNE$} (WFacc);
\draw[arrow, bend left = 25] (WFminDNE) to node[below, text=red]{$\mpe$} (WFseq);
\draw[arrow, bend right = 45] (WFminCor) to node[right, text=red]{$\corDNE$} (WFCor);
\draw[arrow, bend right = 45] (WFaccM) to node[right, text=red]{$\accDNE$} (WFacc);
\draw[arrow] (WFacc) to node[below, text=red]{$\rdec$} (WFseq);
\draw[arrow, bend left=25] (WFminCor) to node[below, text=red]{$\accCor$} (WFaccM);

@@

Other observations that can be gleaned from the figure include the following.
\begin{enumerate}
  \item The implication from $\WFacc$ to $\WFseq$ requires the underlying relation to be decidable.
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item $\WFmin$ is a very strong assumption and implies $\WFseq$ unconditionally.
  The more reasonable $\WFminDNE$ requires the image of every sequence to be $\lnot\lnot$-closed
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (if one assumes that equality on $A$ is decidable).
  \item On the $\lnot\lnot$-side, $\WFminm$ and $\WFminDNEm$ are equivalent. Both are directly implied by $\WFaccm$,
    and all three directly imply $\WFseqm$.
  \item All definitions imply $\WFcorm$.  This weakest definition implies $\WFaccm$ assuming the
    ``constructive contrapositive of accessibility'', $\accCor$.
    %While this condition is implied by $\isWFseqm$, it does not appear to imply anything on its own.

    With this assumption, $\WFacc$ can also be derived from $\WFcor$. % , the stronger version of $\WFcorm$.

@@

  \item If the relation is finitely branching ($\FB$), then $\WFminDNEm$
    implies $\WFaccm$ directly.

@@

We also did not include the inductive definition, Definition \ref{def:WFproperties}.\ref{def:WFind}.
  Logically, it is trivially equivalent to the accessible one, but its Agda formalization has a defect:
  the inductive well-foundedness property lands in the larger universe $\mathtt{Set}_1$, due to
  quantification over all predicates on $A$.  This defect is also present in $\WFmin$, $\WFminDNE$,
  and $\WFcor$.

@@

\subsection{Well-foundedness and rewriting}

One may wonder whether the difference between the various definitions above matters for 
rewriting theory.  In particular, how often are the properties listed in 
Figure \ref{tab:cprop} actually valid for a given rewrite system, so that its termination 
($\WFacc$) could be inferred from a weaker statement, such as $\WFminDNEm$?

On the one hand, \cite{Berardi} proves a general result to the effect that
$\accDNE$ is valid \emph{metatheoretically} for a large class of definable relations.
Specifically, if $R$ is a primitive recursive predicate on the natural numbers and
$\WFaccm(R)$ can be proved in higher-order arithmetic, then an intuitionistic proof 
of $\WFacc(R)$ is also possible.
% situation of separating these notions is complicated by the fact that, for all
% \emph{definable} predicates, the classical and constructive notions do coincide.
% (We have not proved this, but see \cite{Berardi} for a related result.)
So there is little hope of distinguishing between these notions via an explicit counterexample.

On the other hand, some of these properties could be discriminated
by showing that an implication between them yields a ``constructive taboo'':
a proposition known to be unprovable without additional classical assumptions.
For example, the archetype of a well-founded relation is the strict order \verb|<|
on the natural numbers.  The usual induction readily validates $\WFacc$ for \verb|<|.
  At the same time, the file \texttt{WFCounters.agda} shows that
asserting $\WFmin$ for \verb|<| implies that every subset of the natural numbers is decidable.
Since this is clearly not provable in Agda without classical axioms,
this argument shows that the implication from $\WFacc$ to $\WFmin$ is not provable either.

Such distinctions have consequences for rewriting theory.
For example, recall that $x\in \SN_R$ iff $x$ is $\bar{R}$-accessible, where $\bar{R}$ 
is the converse of $R$.  Thus, if $\bar{R}$ is well-founded, $R \models \gSN$.
Yet actually computing the normal form of a given $x$ seems to require a
strong form of decidability of being $R$-minimal, namely:
\[
\tag{$\isMinDec$} \forall x. \left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)
\]
This is weaker than deciding whether $x$ is a normal form:
% Note that deciding whether $x$ is a normal form constitutes the weaker disjunction
$\lnot\lnot\left(\exists y. Ryx\right) \lor \left(\forall y. \lnot Ryx\right)$.

@@

Combined with the natural numbers example in \texttt{WFCounters.agda}, this shows
that the intuitive implication $\SN_R \to \WN_R$ is not actually intuitionistically valid.

% \greyout{From the perspective of rewriting theory, it would seem wild indeed if
% $\SN$ did not imply $\WN$.
%
% These questions motivated us to take a closer look at the issue of well-foundedness.
% Our analysis revealed a number of additional related hypotheses, having a complex
% network of relations.  %These are displayed in Figure \ref{f:wf}.
% These are recorded in the file \verb|Wellfounded.agda| /sacomment{Which file do we now want to refer to here. Or just refer to the entire subfolder?}
%
